\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}

### 2021-11-16 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/04-Longitudinal analysis with GAMs.Rmd
Updated sections

The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors that appear in Figure 3C in [@vishwanath2009]. In the paper, diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10). In the "Treatment" group (chemotherapy) an increase in $\mbox{StO}_2$ is observed through time, while a decrease is seen in the "Control" (saline) group. Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (matching the SD in the original paper). 
The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot)A and the inset, respectively.

@@

#plotting simulated data
f2<-ggplot(dat_sim, aes(x = Day, y = StO2_sim, color = Group, group=subject)) +
    geom_point(show.legend=FALSE,size=1.5,alpha=0.6)+
    geom_line(size=1.1, alpha=0.6)+
    # stat_summary(aes(y = StO2_sim,
    #                  group=Group), 
    #              fun=mean, geom="line",size=1,show.legend = FALSE)+
  labs(y=expression(atop(StO[2],'(simulated)')))+
  theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
  thm1+
    scale_x_continuous(breaks=c(0,2,5,7,10))

@@

#plot smooths and confidence interval for GAM
f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=0.3,
              data=gam_predict,
            show.legend=FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=gam_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'complete')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
  thm1

#plot linear fit for rm-ANOVA
f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),fill=Group),
              alpha=0.3,
              data=lm_predict,
              show.legend = FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=lm_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(paste('StO'[2],' (simulated)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
  thm1

@@

f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=0.3,
              data=m_predict,
            show.legend=FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=m_predict,
              show.legend = TRUE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'missing')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
  thm1

@@

(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors. A: Simulated data that follows previously reported trends (inset) in tumors under chemotherapy (Treatment) or saline (Control) treatment. Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point. Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment. Lines represent trends for each group, shaded regions are 95% confidence intervals. C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time. D: Smooths for the GAM model for the simulated data with 40% of its observations missing. Lines represent trends for each group, shaded regions are 95% empirical Bayesian confidence intervals.

```{r,sim-smooth-plot,fig.width=10, fig.height=7, out.width='75%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}

mult_plot<-f2+inset_element(
  f1, left = 0.01, 
  bottom = 0.5, 
  right = 0.5, 
  top = 1.0)+
  f3+f4+f6+
   plot_annotation(tag_levels='A')&
   ylim(c(-7,75)) &
  theme(
     text=element_text(size=18)
     )&
  thm1

>>>>

The simulated data is based on the reported longitudinal changes in oxygen saturation ($\mbox{StO}_2$) in subcutaneous tumors (Figure 3C in Vishwanath et. al.[@vishwanath2009]). In the paper, diffuse reflectance spectroscopy was used to quantify $\mbox{StO}_2$ changes in both groups at the same time points (days 0, 2, 5, 7 and 10). In the "Treatment" group (chemotherapy) an increase in $\mbox{StO}_2$ is observed through time, while a decrease is seen in the "Control" (saline) group. Following the reported trend, we simulated 10 normally distributed observations at each time point with a standard deviation (SD) of 10% (matching the SD in the original paper). 
The simulated and real data appear in Figure \@ref(fig:sim-smooth-plot),A.

@@

#plotting simulated data
f2<-ggplot(dat_sim, aes(x = Day, 
                        y = StO2_sim, 
                        color = Group, 
                        group=subject)) +
    geom_point(show.legend=FALSE,
               size=1.5,
               alpha=0.6)+
    geom_line(size=0.6, alpha=0.4)+
    geom_line(aes(x = Day, 
               y = StO2,
               color=Group),
           size=1.5,
           data=dat,
           inherit.aes=FALSE,
              show.legend = FALSE)+
  labs(y=expression(atop(StO[2],'(simulated)')))+
  theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )+
  thm1+
    scale_x_continuous(breaks=c(0,2,5,7,10))

@@

#plot smooths and confidence interval for GAM
f3<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=0.5,
              data=gam_predict,
            show.legend=FALSE,
            inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=gam_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'complete')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )

#plot linear fit for rm-ANOVA
f4<-ggplot(data=dat_sim, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),fill=Group),
              alpha=0.5,
              data=lm_predict,
              show.legend = FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=lm_predict,
              show.legend = FALSE)+
  #facet_wrap(~Group)+
  labs(y=expression(paste('StO'[2],' (simulated)')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )

@@

f6<-ggplot(data=dat_missing, aes(x=Day, y=StO2_sim, group=Group)) +
    geom_point(aes(color=Group),size=1.5,alpha=0.5,show.legend = FALSE)+
  geom_ribbon(aes( x=Day,ymin=(fit - 2*se.fit), 
                   ymax=(fit + 2*se.fit),
                   fill=Group
                   ),
              alpha=0.5,
              data=m_predict,
            show.legend=FALSE,
                inherit.aes=FALSE) +
  geom_line(aes(y=fit,
                color=Group),
              size=1,data=m_predict,
              show.legend = TRUE)+
  #facet_wrap(~Group)+
  labs(y=expression(atop(StO[2],'missing')))+
    scale_x_continuous(breaks=c(0,2,5,7,10))+
      theme_classic()+
  theme(
    axis.text=element_text(size=22)
  )

@@

(ref:sim-plot-caption) Simulated data and smooths for oxygen saturation in tumors. A: Simulated data (thin lines) that follows previously reported trends (thick lines) in tumors under chemotherapy (Treatment) or saline (Control) treatment. Simulated data is from  a normal distribution with standard deviation of 10% with 10 observations per time point. Lines indicate mean oxygen saturation B: Smooths from the GAM model for the full simulated data with interaction of Group and Treatment. Lines represent trends for each group, shaded regions are 95% confidence intervals. C: The rm-ANOVA model for the simulated data, which does not capture the changes in each group over time. D: Smooths for the GAM model for the simulated data with 40% of its observations missing. Lines represent trends for each group, shaded regions are 95% empirical Bayesian confidence intervals.

```{r,sim-smooth-plot,fig.width=12, fig.height=7, out.width='100%', fig.align='center',echo=FALSE,message=FALSE, warning=FALSE,fig.cap='(ref:sim-plot-caption)'}

mult_plot<-f2+f3+f4+f6+
   plot_annotation(tag_levels='A')&
   ylim(c(-7,75)) &
  theme(
     text=element_text(size=18)
     )

### 2021-11-16 Manuscripts/Manuscript_by_chapters-SIM_Revisions/sections/05-Discussion.Rmd
updated GAM section
Updated sections

This "model misspecification" error, also is known as a "Type III" error [@dennis2019] is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model. We acknowledge that polynomial effects can be used in LMEMs, and that it could be argued that in Section \@ref(simulation) a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects causes more questions than answers in this case (i.e., what is the appropriate polynomial effect to use? quadratic?, cubic?, what is the biological interpretation of a polynomial effect?). Moreover, polynomial effects have other associated issues that we have described in Section {#GAM-theory}.

>>>>

This "model misspecification" error, also is known as a "Type III" error [@dennis2019] is particularly important because although the _p-value_ is the common measure of statistical significance, the validity of its interpretation is determined by the agreement of the data and the model. Polynomial effects can be used in LMEMs, and it could be argued that in Section \@ref(simulation) a LMEM with quadratic effects could have been fitted, but in reality the true function in the data is not known; using polynomial effects presents more questions than answers in this case (i.e., what is the appropriate polynomial effect to use? quadratic?, cubic?, what is the biological interpretation of a polynomial effect?). Moreover, polynomial effects have other associated issues that we have described in Section \@ref(GAM-theory).
